---
title: "Simulate genetic effects and PRS"
output: html 
---


# Genetic and environmental effects

Simulate parental genetic and and environmental effects for given heritability ($h^2$) and phenotypic variance $V_P$

```{r}
set.seed(14) #lucky number
h2 = 0.8
V_P = 1
V_Tagged = 0.45 #Yengo et al., 2018 states that several previous gwas have found snp heritability to be 45%
V_nTagged = V_A - V_Tagged
V_A = h2*V_P
V_E = (1-h2)*V_P
n = 10000

#Creating a genetic variable that accounts for all genetic variance
maternal_genetic = rnorm(n, mean=0, sd = sqrt(V_A))
paternal_genetic = rnorm(n, mean=0, sd = sqrt(V_A))

#Creating a genetic variable that captures ALL tagged genetic variance
maternal_genetic_tagged = rnorm(n, mean=0, sqrt(V_Tagged))
paternal_genetic_tagged = rnorm(n, mean=0, sqrt(V_Tagged))

#Creating a genetic variable that captures ALL non-tagged genetic variance
maternal_genetic_ntagged = rnorm(n, mean=0, sqrt(V_nTagged))
paternal_genetic_ntagged = rnorm(n, mean=0, sqrt(V_nTagged))


maternal_environment = rnorm(n, mean=0, sd = sqrt(V_E))
paternal_environment = rnorm(n, mean=0, sd = sqrt(V_E))

maternal_phenotype = maternal_genetic + maternal_environment
paternal_phenotype = paternal_genetic + paternal_environment


```

Offspring genetic is the average of the parents plus segregation variance $V_\mathrm{seg} = \frac{1}{2}V_A$

```{r}

offspring_segregation = rnorm(n, 0, sd=sqrt(V_A/2))
offspring_segregation_tagged = rnorm(n, 0, sd=sqrt(V_Tagged/2))

offspring_genetic = (maternal_genetic + paternal_genetic)/2 + offspring_segregation

offspring_environment = rnorm(n, mean=0, sd = sqrt(V_E))

offspring_phenotype = offspring_genetic + offspring_environment


```

Check offspring on parent regression

```{r}

2*coef(lm(offspring_phenotype ~ maternal_phenotype))[[2]]

```

```{r}

2*coef(lm(offspring_phenotype ~ paternal_phenotype))[[2]]

```

```{r}
coef(lm(offspring_phenotype ~ I((maternal_phenotype + paternal_phenotype)/2)))[[2]]
```

# PRS

Create *noisy* PRS. PRS is the parental genetic effect plus a random deviation that is `prs_scale` times the genetic effects, then turn them back into standardized variables
```{r}
 
prs_scale <- 10
maternal_prs = scale(rnorm(n, mean=maternal_genetic_tagged, sd=sqrt(prs_scale*V_Tagged)))[,1]
paternal_prs = scale(rnorm(n, mean=paternal_genetic_tagged, sd=sqrt(prs_scale*V_Tagged)))[,1]
```


Offspring PRS is the average of their parents' PRSs plus a noisy deviation based on the offspring's own segregation deviation
```{r}
offspring_prs = scale((maternal_prs + paternal_prs)/2 + rnorm(n, mean=offspring_segregation_tagged, sd=sqrt(prs_scale*V_Tagged))/4)[,1]
```
  

Check that expected parent-offspring PRS correlation is about 0.5

```{r}
cor(maternal_prs, offspring_prs)
cor(paternal_prs, offspring_prs)
```

# Simulated data

Put all values into a data frame.
```{r}

cohort <- data.frame(maternal_genetic, maternal_environment, maternal_phenotype, maternal_prs,
                 paternal_genetic, paternal_environment, paternal_phenotype, paternal_prs,
                 offspring_genetic, offspring_environment, offspring_phenotype, offspring_prs)
write.table(cohort, "cohort", row.names=F, quote=F)
```

# Regression models

Regress offspring phenotypes on genetic values
```{r}
summary(lm(offspring_phenotype ~ offspring_genetic + maternal_genetic + paternal_genetic, data=cohort))
```

Regress offspring phenotypes on PRS values
```{r}
summary(lm(offspring_phenotype ~ offspring_prs, data=cohort))
summary(lm(offspring_phenotype ~ offspring_prs + maternal_prs + paternal_prs, data=cohort))
summary(lm(offspring_phenotype ~ offspring_prs + maternal_prs + paternal_prs + maternal_phenotype + paternal_phenotype, data=cohort))

```

Pathway models using PRSs  

Simple model 
```{r}

library(lavaan)

data <- read.table("cohort", header=T)
data <- data[complete.cases(data),]

sink("hei_pa_sim_m1_prs.txt")

# Define model

model <- ' 

#direct effects

offspring_phenotype ~ c*maternal_prs
offspring_phenotype ~ f*paternal_prs            

#mediator

offspring_prs ~ 0.5*maternal_prs
offspring_prs ~ 0.5*paternal_prs

offspring_phenotype ~ b*offspring_prs

# indirect effects (a*b, d*b)

ab := 0.5*b
db := 0.5*b

# total effects 

total_mum := c + (0.5*b)
total_dad := f + (0.5*b)

# covariance

maternal_prs ~~ paternal_prs
'

# Fit the model
fitM <- sem(model, data=data, estimator='ML')
print(summary(fitM, standardized=TRUE, fit.measures=TRUE))
print(AIC(fitM))
sink()

```

Extended model
```{r}

rm(list=ls())

library(lavaan)

data <- read.table("cohort", header=T)
data <- data[complete.cases(data),]

sink("hei_pa_sim_m2_prs.txt") #adjusted extended model#

# Define model

model <- '

#direct effects

offspring_phenotype ~ c*maternal_prs
offspring_phenotype ~ f*paternal_prs        

#mediator

offspring_prs ~ 0.5*maternal_prs
offspring_prs ~ 0.5*paternal_prs

offspring_phenotype ~ b*offspring_prs

paternal_phenotype ~ g*paternal_prs
maternal_phenotype ~ h*maternal_prs

offspring_phenotype ~ i*paternal_phenotype
offspring_phenotype ~ j*maternal_phenotype

# indirect effects (a*b, d*b)

ab := 0.5*b
db := 0.5*b
gi := g*i
hj := h*j


# total effects 

total_mum := c + (0.5*b) + (h*j) 
total_dad := f + (0.5*b) + (g*i) 

# covariance

maternal_prs ~~ paternal_prs
maternal_phenotype ~~ paternal_phenotype

'

# Fit the model
fitM <- sem(model, data=data, estimator='ML')
print(summary(fitM, standardized=TRUE, fit.measures=TRUE))
print(AIC(fitM))
sink()

```

Extended Restricted Model
```{r}
#############################
#EXTENDED (RESTRICTED) MODEL#
#############################

rm(list=ls())

library(lavaan)

data <- read.table("cohort", header=T)
data <- data[complete.cases(data),]

sink("hei_pa_sim_m3_prs.txt") #adjusted extended model#

# Define model

model <- ' 

#direct effects

#mediator

offspring_prs ~ 0.5*maternal_prs
offspring_prs ~ 0.5*paternal_prs

offspring_phenotype ~ b*offspring_prs

paternal_phenotype ~ g*paternal_prs
maternal_phenotype ~ h*maternal_prs

offspring_phenotype ~ i*paternal_phenotype
offspring_phenotype ~ j*maternal_phenotype

# indirect effects (a*b, d*b)

ab := 0.5*b
db := 0.5*b
gi := g*i
hj := h*j


# total effects 

total_mum := (0.5*b) + (h*j) ###### THIS IS WHAT I HAVE ADDED - NOT SURE IF ITS RIGHT THO #####
total_dad := (0.5*b) + (g*i) ###### THIS IS WHAT I HAVE ADDED - NOT SURE IF ITS RIGHT THO #####

# covariance

maternal_prs ~~ paternal_prs
maternal_phenotype ~~ paternal_phenotype
'

# Fit the model
fitM <- sem(model, data=data, estimator='ML')
print(summary(fitM, standardized=TRUE), fit.measures=TRUE)
print(AIC(fitM))
sink()

```


Pathway models using genetic variable  

Simple Model
```{r}

library(lavaan)

data <- read.table("cohort", header=T)
data <- data[complete.cases(data),]

sink("hei_pa_sim_m1_genetic.txt")

# Define model

model <- ' 

#direct effects

offspring_phenotype ~ c*maternal_genetic
offspring_phenotype ~ f*paternal_genetic            

#mediator

offspring_genetic ~ 0.5*maternal_genetic
offspring_genetic ~ 0.5*paternal_genetic

offspring_phenotype ~ b*offspring_genetic

# indirect effects (a*b, d*b)

ab := 0.5*b
db := 0.5*b

# total effects 
total_mum := c + (0.5*b)
total_dad := f + (0.5*b)

# covariance

maternal_genetic ~~ paternal_genetic
'

# Fit the model
fitM <- sem(model, data=data, estimator='ML')
print(summary(fitM, standardized=TRUE, fit.measures=TRUE))
print(AIC(fitM))
sink()
```

Extended Model
```{r}

rm(list=ls())

library(lavaan)

data <- read.table("cohort", header=T)
data <- data[complete.cases(data),]

sink("hei_pa_sim_m2_genetic.txt") #adjusted extended model#

# Define model

model <- '

#direct effects

offspring_phenotype ~ c*maternal_genetic
offspring_phenotype ~ f*paternal_genetic        

#mediator

offspring_genetic~ 0.5*maternal_genetic
offspring_genetic ~ 0.5*paternal_genetic

offspring_phenotype ~ b*offspring_genetic

paternal_phenotype ~ g*paternal_genetic
maternal_phenotype ~ h*maternal_genetic

offspring_phenotype ~ i*paternal_phenotype
offspring_phenotype ~ j*maternal_phenotype

# indirect effects (a*b, d*b)

ab := 0.5*b
db := 0.5*b
gi := g*i
hj := h*j


# total effects 

total_mum := c + (0.5*b) + (h*j) ###### THIS IS WHAT I HAVE ADDED - NOT SURE IF ITS RIGHT THO #####
total_dad := f + (0.5*b) + (g*i) ###### THIS IS WHAT I HAVE ADDED - NOT SURE IF ITS RIGHT THO #####

# covariance

maternal_genetic ~~ paternal_genetic
maternal_phenotype ~~ paternal_phenotype

'

# Fit the model
fitM <- sem(model, data=data, estimator='ML')
print(summary(fitM, standardized=TRUE, fit.measures=TRUE))
print(AIC(fitM))
sink()
```

Extended (Restricted) Model
```{r}

rm(list=ls())

library(lavaan)

data <- read.table("cohort", header=T)
data <- data[complete.cases(data),]

sink("hei_pa_sim_m3_genetic.txt") #adjusted extended model#

# Define model

model <- ' 

#direct effects

#mediator
offspring_genetic ~ 0.5*maternal_genetic
offspring_genetic ~ 0.5*paternal_genetic

offspring_phenotype ~ b*offspring_genetic

paternal_phenotype ~ g*paternal_genetic
maternal_phenotype ~ h*maternal_genetic

offspring_phenotype ~ i*paternal_phenotype
offspring_phenotype ~ j*maternal_phenotype

# indirect effects (a*b, d*b)
ab := 0.5*b
db := 0.5*b
gi := g*i
hj := h*j


# total effects 

total_mum := (0.5*b) + (h*j) ###### THIS IS WHAT I HAVE ADDED - NOT SURE IF ITS RIGHT THO #####
total_dad := (0.5*b) + (g*i) ###### THIS IS WHAT I HAVE ADDED - NOT SURE IF ITS RIGHT THO #####

# covariance

maternal_genetic ~~ paternal_genetic
maternal_phenotype ~~ paternal_phenotype
'

# Fit the model
fitM <- sem(model, data=data, estimator='ML')
print(summary(fitM, standardized=TRUE), fit.measures=TRUE)
print(AIC(fitM))
sink()

```








